Fractional Types

### 2013-02-17 frac.tex
sec 3 kind of ok

The language {{Pi}} models isomorphisms of values rather well.  However,
purposefully, it has no distinguished notion of \emph{input} or
\emph{output}; more precisely, because of strict preservation of information,
these two concepts coincide.  This is the root of reversibility. But if we
want to model functions of any flavor as values, we need to differentiate
between these notions.  Our idea, inspired by computational dualities and
polarity, is to introduce a \emph{formal dual} for our types.  We thus
consider a type in a negative position to possess negative information, as it
is really a request for information.  Since information is logarithmic, this
means that a request for information should behave like a \emph{fractional
  type}.

@@

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.  However,
we think of a value like {{1/left ()}} as a \emph{first-class constraint}
that can only be satisfied if it is matched with an actual value {{left ()}}
or as representing the absence of some information (i.e., \emph{negative
  information}), that can only be supplied by {{left ()}}. The combinator
{{eta*_b : 1 <-> 1/b * b}} represents a fission point which creates --- out
of no information --- an equal amount of negative and positive
information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
up an equal amount of negative and positive information producing no residual
information.  The best analogy is probably that of vectors and co-vectors,
also know as column vectors and row vectors (respectively): in the finite
dimensional case, they are isomorphic, but they are nevertheless quite
different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words,
since we are explicitly interested in \emph{isomorphisms}, we should not
conflate negative-information values with positive-information values just
because they happen to be isomorphic. (See Sec.~\ref{sec:cat} for a more
thorough discussion about the meaning of fractional types and values in an
algebraic and categorical sense.)

For now, we proceed with an operational understanding of the language. The
semantics we provide is simple: each combinator {{c : b1 <-> b2}} is
interpreted as a \emph{relation} between the sets denoted by {{b1}} and
{{b2}} respectively.

@@

\newpage
~
\newpage

\roshan{In the judgments below it is misleading that only {{eta*}}
  appears indexed by a type. To construct each family of relations,
  the corresponding definition must be indexed by their types. In
  other words, there is nothing special about {{eta*}}. Further, I am
  unsure of how to format this so that the combinators are correctly
  indexed by their types for both primitives and for composition
  combinators. We could do the following, but it seems to be a
  overkill.}

%subcode{proof} include main
%@ v1 in [[b1]]
%@ v2 in [[b2]]
%@ v3 in [[b3]]
%@@ (v1, (v2, v3)) ~assocl*~ ((v1, v2), v3)

In the previous section, we have been able to maintain the simple idea that a
syntactic isomorphism denotes a \emph{bijection}. In richer languages, e.g.,
in a Turing complete language that admits non-termination, isomorphisms
necessarily denote \emph{partial
  bijections}~\cite{rc2011,James:2012:IE:2103656.2103667}. As we demonstrate
in this section, once we make the extension to higher-order functions, we
must interpret the syntactic isomorphisms as \emph{relations}. The intuitive
idea, is that the fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be
able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
undefined, i.e., it does not produce {{():1}}. In other words, some of the
combinators are not partial and some produce more than one result for a
particular input, and hence we move to the setting of general binary
relations to interpret our combinators.

This setting is not only consistent with logical reversibility but is
arguably the \emph{right} framework for considering expressive reversible
languages (including the interesting examples of quantum languages).  Indeed,
even though the categorical semantics of {{PiTF}} is not the main focus
of this paper, we find it helpful to state that the category of finite sets
and relations in which we are going to interpret our language, is an example
of a \emph{biproduct dagger compact closed category} which are categories
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} Two of the most relevant features of
these categories for our purposes are that they support an adjoint for each
morphism (to model reversibility) and that they support higher-order relation
(built from the categorical dagger which is modeled by our fractional
types). The latter is intuitively clear as any relation between $A$ and $B$
can be represented as a subset of {{A * B}}, i.e., as a value.

To interpret {{PiTF}} in this category, we associate a finite set
{{ [[ b ]] }} with each type {{b}} and a relation {{ [[ c ]] }} over
{{ [[ b1 ]] }} and {{ [[ b2 ]] }} for each combinator {{ c : b1 <-> b2 }}.

The language {{PiTF}} is still logically reversible. This is intuitively
clear as the relation denoted by a combinator can be used in the reverse
direction by swapping inputs and outputs.

\roshan{ In the rules above it is unclear as to which v is chosen in
  the rule for eta. It is unclear that these rules are logically
  reversible.  }

\begin{proposition}[Logical Reversibility]
\label{chx:prop:logrev-tracep}
For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
{{v2 : b2}} we have {{v1 ~~c~~ v2}} iff {{v2 ~~(sym c)~~ v1}}.
\end{proposition}

Some thoughts on a relational semantics. We define {{ a <-> b }}
$\equiv$ {{a R b}}

\noindent
We now have primitives:
%subcode{opsem} include main
% swap*:b1*b2<->b2*b1 '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
% swap+:b1+b2<->b2+b1 '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) 
%                     && ~~~| v1 in b1, v2 in b2]}
% assocl*:b1*(b2*b3)<->(b1*b2)*b3 '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) 
%                   & & ~~~|~ v1 in b1, v2 in b2, v3 in b3]}
etc..

\noindent
closure combinators:
%subcode{opsem} include main
% sym c '= {[(v2, v1) ~|~ (v1, v2) in c ]}
% c1 (;) c2 '= {[(v1, v3) ~|~ v2 in b2, (v1, v2) in c1, (v2, v3) in c2 ]}
% c1 (*) c2 '= {[((v1, v2), (v3, v4)) ~|~ (v1, v3) in c1, (v2, v4) in c2 ]}
etc..

\noindent
Now we have 
%subcode{opsem} include main
% eta* '= {[ ((), (v, 1/v)) ~|~ v in b ]}
% eps* '= {[ ((v, 1/v), ()) ~|~ v in b ]}

Now {{PiTF}} is logically reversible in that for all {{c:b1<->b2}} we
have {{c^{dagger}: b2 <-> b1}} such that 
%
{{c^{dagger} = {[(v2, v1)~|~(v1,v2) in c ]} }}. This is
straightforward from the definitions above.

I think defining eval is an open problem.  Consider a few relations:

\begin{enumerate}
\item 
{{empty = {[ ]} }}

\item 
{{full =  {[ R, G, B ]} * {[ R, G, B ]} }}

\item 
{{id  =  {[ (R, R), (G, G), (B, B) ]} }}

\item 
{{cycle = {[ (R, R), (R, G), (G, G), (G, B), (B, B), (B, R) ]} }}
\end{enumerate}

Can we define an {{eval}} that meaningfully distinguishes these? What
can we say about {{(c (;) c^^^) v}} based on {{eval}}?  Here are
several possible {{eval}} definitions:

\begin{enumerate}
\item {{ c v = v'}} where {{ v' in {[ v' ~|~ (v, v') in c ]} }}. The
  eval maps {{b1 -> b2}}.

\item {{ c v = {[ v' ~|~ (v, v') in c ]} }}. The eval maps {{b1 -> Pow(b2)}}.

\item {{ c vs = {[ v' ~|~ (v, v') in c, v in vs ]} }}. where
  {{vs:Pow(b1)}}. The eval maps {{Pow(b1) -> Pow(b2)}}.

\end{enumerate}

For the relations above, none of the suggested eval definitions give a
logically reversible semantics.

\roshan{If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
{{b2}}? If yes, then we have a model for {{PiTF}} where the objects
are sets {{b}} and the morphisms are the relations between objects of
equal size.}

\roshan{ I suspect that all relations can be constructed, but we have
  to show this. This is not obvious.}

>>>>

The language {{Pi}} models isomorphisms of values rather well.  However,
purposefully, it has no distinguished notion of \emph{input} or
\emph{output}; more precisely, because of strict preservation of information,
these two concepts coincide.  This is the root of reversibility. But if we
want to model functions of any flavor as values, we need to differentiate
between these notions.  Our idea, inspired by computational
dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
polarity~\cite{Girard87tcs,10.1109/LICS.2010.23}, and the categorical notion
of \emph{compact
  closure}~\cite{Selinger:2007:DCC:1229185.1229207,Abramsky:2004:CSQ:1018438.1021878},
is to introduce a \emph{formal dual} for our types.  We thus consider a type
in a negative position to possess negative information, as it is really a
request for information.  Since information is logarithmic, this means that a
request for information should behave like a \emph{fractional type}.

@@

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.  However,
we think of a value like {{1/left ()}} as a \emph{first-class constraint}
that can only be satisfied if it is matched with an actual value {{left ()}}
or as representing the absence of some information (i.e., \emph{negative
  information}), that can only be reconciliated by {{left ()}}. The
combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission point which
creates --- out of no information --- an equal amount of negative and
positive information. Symmetrically, the combinator 
{{eps*_b : 1/b * b <-> 1}} 
matches up an equal amount of negative and positive information producing no
residual information.  The best analogy is probably that of vectors and
co-vectors, also know as column vectors and row vectors (respectively) where
the multiplication of the row vector by the column vector is a scalar. In the
finite dimensional case, they are isomorphic, but they are nevertheless quite
different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words,
since we are explicitly interested in \emph{isomorphisms}, we should not
conflate negative-information values with positive-information values just
because they happen to be isomorphic. (See Sec.~\ref{sec:cat} for a more
thorough discussion about the meaning of fractional types and values in both an
algebraic and categorical sense.) For now, we proceed with an operational 
understanding of the language.

The semantics we provide is simple: each combinator {{c : b1 <-> b2}} is
interpreted as a \emph{relation} between the sets denoted by {{b1}} and
{{b2}} respectively.

@@

If it were not for the cases of {{eta*}} and {{eps*}}, each combinator 
{{c : b1 <-> b2}} would define a total one-to-one function between the sets 
{{ [[ b1 ]] }} and {{ [[ b2 ]] }}. 
This is indeed the situation for the language {{Pi}} 
defined in the previous section. In the presence of fractionals, this simple 
interpretation of combinators as one-to-one functions is not sufficient.
\footnote{The interpretation of combinators as one-to-one functions is also 
not valid in the presence of recursive 
types~\cite{rc2011,James:2012:IE:2103656.2103667}.}
The intuitive reason, is that the fission point combinator 
{{eta*_b : 1 <-> 1/b * b}} must be
able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
undefined, i.e., it does not produce {{():1}}.

The relational semantics can be given an operational interpretation by
defining two functions: a forward evaluator and a backwards evaluator that
exercise the relation in opposite directions. It follows that if the forward
evaluation of {{c v1}} produces {{v2}} as a possible answer then the
backwards evaluation of {{c v2}} produces {{v1}} as a possible answer. More
generally, the semantics of {{PiTF}} using sets and relations, i.e., in the
category of finite sets and relations suggests a few additional immediate
properties.  The category of finite sets and relations is an example of a
\emph{biproduct dagger compact closed category} which are the categories
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} Two of the most relevant features of
these categories for our purposes are that they support an adjoint for each
morphism (to model reversibility) and that they support higher-order relation
(built from the categorical dagger which is modeled by our fractional
types). The latter is intuitively clear as any relation between $A$ and $B$
can be represented as a subset of {{A * B}}, i.e., as a value. We return to
such properties of {{PiTF}} in Sec.~\ref{sec:cat} after we present
programming examples that illustrate {{PiTF}} and its ability to express
various higher-order notions.

% \roshan{ In the rules above it is unclear as to which v is chosen in
%   the rule for eta. It is unclear that these rules are logically
%   reversible.  }

% \begin{proposition}[Logical Reversibility]
% \label{chx:prop:logrev-tracep}
% For all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and 
% {{v2 : b2}} we have {{v1 ~~c~~ v2}} iff {{v2 ~~(sym c)~~ v1}}.
% \end{proposition}

% Some thoughts on a relational semantics. We define {{ a <-> b }}
% $\equiv$ {{a R b}}

% \noindent
% We now have primitives:
% %subcode{opsem} include main
% % swap*:b1*b2<->b2*b1 '= {[ ((v1, v2), (v2, v1)) ~|~ v1 in b1, v2 in b2 ]}
% % swap+:b1+b2<->b2+b1 '= {[ ((left v1), (right v1)) and ((right v2), (left v2)) 
% %                     && ~~~| v1 in b1, v2 in b2]}
% % assocl*:b1*(b2*b3)<->(b1*b2)*b3 '= {[ ((v1, (v2, v3)), ((v1, v2), v3)) 
% %                   & & ~~~|~ v1 in b1, v2 in b2, v3 in b3]}
% etc..

% \noindent
% closure combinators:
% %subcode{opsem} include main
% % sym c '= {[(v2, v1) ~|~ (v1, v2) in c ]}
% % c1 (;) c2 '= {[(v1, v3) ~|~ v2 in b2, (v1, v2) in c1, (v2, v3) in c2 ]}
% % c1 (*) c2 '= {[((v1, v2), (v3, v4)) ~|~ (v1, v3) in c1, (v2, v4) in c2 ]}
% etc..

% \noindent
% Now we have 
% %subcode{opsem} include main
% % eta* '= {[ ((), (v, 1/v)) ~|~ v in b ]}
% % eps* '= {[ ((v, 1/v), ()) ~|~ v in b ]}

% Now {{PiTF}} is logically reversible in that for all {{c:b1<->b2}} we
% have {{c^{dagger}: b2 <-> b1}} such that 
% %
% {{c^{dagger} = {[(v2, v1)~|~(v1,v2) in c ]} }}. This is
% straightforward from the definitions above.

% I think defining eval is an open problem.  Consider a few relations:

% \begin{enumerate}
% \item 
% {{empty = {[ ]} }}

% \item 
% {{full =  {[ R, G, B ]} * {[ R, G, B ]} }}

% \item 
% {{id  =  {[ (R, R), (G, G), (B, B) ]} }}

% \item 
% {{cycle = {[ (R, R), (R, G), (G, G), (G, B), (B, B), (B, R) ]} }}
% \end{enumerate}

% Can we define an {{eval}} that meaningfully distinguishes these? What
% can we say about {{(c (;) c^^^) v}} based on {{eval}}?  Here are
% several possible {{eval}} definitions:

% \begin{enumerate}
% \item {{ c v = v'}} where {{ v' in {[ v' ~|~ (v, v') in c ]} }}. The
%   eval maps {{b1 -> b2}}.

% \item {{ c v = {[ v' ~|~ (v, v') in c ]} }}. The eval maps {{b1 -> Pow(b2)}}.

% \item {{ c vs = {[ v' ~|~ (v, v') in c, v in vs ]} }}. where
%   {{vs:Pow(b1)}}. The eval maps {{Pow(b1) -> Pow(b2)}}.

% \end{enumerate}

% For the relations above, none of the suggested eval definitions give a
% logically reversible semantics.

% \roshan{If {{|b1|=|b2|}}, can we define all relations between {{b1}} and
% {{b2}}? If yes, then we have a model for {{PiTF}} where the objects
% are sets {{b}} and the morphisms are the relations between objects of
% equal size.}

% \roshan{ I suspect that all relations can be constructed, but we have
%   to show this. This is not obvious.}

% \roshan{In the judgments below it is misleading that only {{eta*}}
%   appears indexed by a type. To construct each family of relations,
%   the corresponding definition must be indexed by their types. In
%   other words, there is nothing special about {{eta*}}. Further, I am
%   unsure of how to format this so that the combinators are correctly
%   indexed by their types for both primitives and for composition
%   combinators. We could do the following, but it seems to be a
%   overkill.}

% %subcode{proof} include main
% %@ v1 in [[b1]]
% %@ v2 in [[b2]]
% %@ v3 in [[b3]]
% %@@ (v1, (v2, v3)) ~assocl*~ ((v1, v2), v3)
